Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-04-30 RNARedPrint-Bioinfo2018.tex
Minor edits in Sec. 4; add todo note

@@

\subsection{Targeting Turner energies and \GCb\%}
We implemented our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for given
target structures and weights $\pi_1,\dots,\pi_n$.
Moreover, we used multi-dimensional Boltzmann sampling, as described in Section~\ref{sec:multiBoltzmann}, to target specific energies and \GCb\%.  Our tool \ourprog{} evaluates energies according to the base
pair energy model or a stacking energy model $\EnergyStacking$, whose parameters 
were fitted to best approximate Turner energies (Supp. Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

To capture the realistic Turner model $\EnergyTurner$ while keeping the computational cost in check, we exploited the excellent linear fit  of $\EnergyTurner$  with the fitted stacking model $\EnergyStacking$ (supp. Fig~\ref{fig:training-cor}). Namely, we observed a structure-specific affine dependency between the Turner and stacking energy models, so that $\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$ for any structure $R$ and sequence $S$. We learned the $(\gamma,\delta)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a predetermined value.  Finally, we adjusted the targeted energies within our stacking model to $\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in order to reach, on average, the right targeted energy $\EnergyTurner^{\star}$ in the Turner model.

Fig.~\ref{fig:energydist} illustrates our above strategy. Fig.~\ref{fig:energydist}A
shows the relative distance between $\EnergyTurner$  and $\EnergyStacking$, observed for instances of the \texttt{Modena} benchmark. For each structure, $n=1\,000$ sequences were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported the difference between the expected and sampled energies for both the Turner model and its approximation using a stacking model. 
Fig.~\ref{fig:energydist}B shows the Turner energy, for a single instance of the 3-structures \texttt{Modena}  benchmark, using different weights benchmark. These results demonstrate that our multidimensionnal Boltzmann sampling strategy can accurately and, to a large extent, independently shift the Turner energies of sampled sequences to prescribed targets.

>>>>

\usepackage{todonotes}

@@

\subsection{Targeting Turner energies and \GCb\%}
We implemented our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
given target structures and weights $\pi_1,\dots,\pi_n$.  Moreover, we
used multi-dimensional Boltzmann sampling, as described in
Section~\ref{sec:multiBoltzmann}, to target specific energies and
\GCb\%.  Our tool \ourprog{} evaluates energies according to the
stacking energy model $\EnergyStacking$, whose parameters were fitted
to best approximate Turner energies. As well, we implemented and
fitted a base pair energy model for \ourprog{}, which was not studied
for its targeting performance (both models:
Supp. Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

To capture the realistic Turner model $\EnergyTurner$ while keeping
the computational cost in check, we exploited the close correlation of
$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
(Supp. Fig~\ref{fig:training-cor}). More precisely, we observed a
structure-specific affine dependency between the Turner and stacking
energy models, so that
$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
for any structure $R$ and sequence $S$. We learned the
$(\gamma,\delta)$ parameters from a set of sequences generated with
homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a
predetermined value.  Finally, we adjusted the targeted energies
within our stacking model to
$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
order to reach, on average, the targeted energy
$\EnergyTurner^{\star}$ in the Turner model.

Fig.~\ref{fig:energydist} illustrates the described
strategy. \todo{Update Figure 2A}Fig.~\ref{fig:energydist}A
illustrates our capability to target Turner energies based on the
stacking energy model. It shows the relative distance between
$\EnergyTurner$ and $\EnergyStacking$, observed for instances of the
\texttt{Modena} benchmark. For each structure, $n=1\,000$ sequences
were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported
the difference between the expected and sampled energies for both the
Turner model and its approximation using a stacking model.  For an
example instance of the \texttt{Modena} benchmark \texttt{3str},
Fig.~\ref{fig:energydist}B shows the Turner energy distributions of
the single structures as they result from sampling with different weight
parameters. The figure illustrates how our multidimensionnal Boltzmann
sampling strategy can accurately and, to a large extent, independently
shift the Turner energies of sampled sequences to prescribed targets.
